Towards a Mechanized Proof of Noninterference

### 2013-12-18 paper/paper.tex
Proofs proofs proofs for the type system.

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{lemma}{Lemma}

@@

    \infer[V-Var]
    {\Gamma(x) = t}
    {\typeRule{\Gamma}{x}{t}}

@@

Before discussing the task of mechanizing a noninterference theorem for
\langName, it is useful to first view what the natural language proof would
look like first.  We first start by proving some supporting lemmas and finish
with a proof of noninterference for the language.  Since \langName is a stripped
down version of \origLang, the proof will look like a simplified version of the
proof found in Pottier and Simonet's paper~\cite{InfoFlowML}.

234101415

@@

\begin{lemma}[Substitution]
  \label{lem:stuck_bracket}
  Assume $\typeRule{ }{v}{s}$.  Then, $\typeRule{\Gamma[x \mapsto s]}{v'}{t}$
  implies $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.  Also $\typeRule{pc,
  \Gamma[x \mapsto s]}{e}{t}$ implies $\typeRule{pc, \Gamma}{e[x \Leftarrow
  v]}{t}$.
\end{lemma}
\begin{proof}
  TODO
\end{proof}

@@

We would like to thank Jon Sailor from Facebook for helping us over the phone
work through some of the more difficult problems we ran into using Coq.

>>>>

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem*{case}{Case}

@@

    \infer[V-Var]
    {t \in \Gamma(x)}
    {\typeRule{\Gamma}{x}{t}}

@@

Before discussing the task of mechanizing a noninterference theorem for
\langName, it is useful to first view what the natural language proof would
look like first.  We first start by proving some supporting lemmas and finish
with a proof of noninterference for the language.  Since \langName is a stripped
down version of \origLang, the proofs will look like a simplified version of the
proofs found in Pottier and Simonet's paper~\cite{InfoFlowML}.

@@

\begin{lemma}
  \label{lem:guard_subtype}
  If $l' \leq l$ and $l \guards t$ and $t \leq t'$, then $l' \guards t'$.
\end{lemma}
\begin{proof}
  $l \guards t$ is equivalent to $l \leq level(t)$ where $level$ is a function
  that takes a type and returns the label on that type.  It's simple to see that
  $level$ is covariant in its argument.  The result follows.
\end{proof}

\begin{lemma}[Weakening]
  \label{lem:weakening}
  $pc' \leq pc$ and $\typeRule{pc, \Gamma}{e}{t}$ imply $\typeRule{pc',
  \Gamma}{e}{t}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\typeRule{pc, \Gamma}{e}{t}$.  By the
  monotonicity of $\sqcup$, rule \textsc{V-Sub}, Lemma~\ref{lem:guard_subtype},
  and the inductive hypothesis, it can be checked that the premise remains valid
  when $pc$ is lowered.
\end{proof}

\begin{lemma}[Projection]
  \label{lem:projection}
  Let $i \in \{1, 2\}$.  If $\typeRule{\Gamma}{v}{t}$ then
  $\typeRule{\Gamma}{\lift{v}_i}{t}$.  If $\typeRule{pc, \Gamma}{e}{t}$ then
  $\typeRule{pc, \Gamma}{\lift{e}_i}{t}$.
\end{lemma}
\begin{proof}
  By induction on the derivation for $v$ and $e$.  The only interesting case is
  that of \textsc{E-Bracket}, since other cases end up either being trivial or
  using the inductive hypothesis for a case of \textsc{E-Bracket}.  The
  expression is $<e_1 | e_2>$.  Since $pc \sqcup pc' \leq pc$, by
  Lemma~\ref{lem:weakening} and the inductive hypothesis, we can prove the
  typing judgements for both branches.
\end{proof}

\begin{lemma}[Guard]
  \label{lem:guard}
  If $\typeRule{\Gamma}{< v_1 | v_2 >}{t}$ then there exists $pc' \in H$ s.t.
  $pc' \guards t$.
\end{lemma}
\begin{proof}
  Through Lemma~\ref{lem:guard_subtype}, we can see that we can always construct
  a typing derivation that does not end with \textsc{V-Sub}.  As a result, the
  derivation can always end with \textsc{V-Bracket}, which has the premises $pc'
  \in H$ and $pc' \guards t$.
\end{proof}

\begin{lemma}[Substitution]
  \label{lem:stuck_bracket}
  Assume $\typeRule{ }{v}{s}$.  Then, $\typeRule{\Gamma[x \mapsto s]}{v'}{t}$
  implies $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.  Also $\typeRule{pc,
  \Gamma[x \mapsto s]}{e}{t}$ implies $\typeRule{pc, \Gamma}{e[x \Leftarrow
  v]}{t}$.
\end{lemma}
\begin{proof}
  We can prove both statements by induction on the typing derivation for $v'$
  and $e$.  In this proof, we focus only on a subset of the cases.

  \begin{case}[\textsc{V-Int}]
    This is trivial, since $x$ does not get substituted in $v'$ and so the
    statement becomes a tautology.
  \end{case}

  \begin{case}[\textsc{V-Var}]
    If $v'$ is $x$, then we are in the case where $s \in t$.  Thus, we can
    conclude that given $\typeRule{ }{v}{s}$, $\typeRule{\Gamma[x \mapsto
    s]}{v'}{t}$ by \textsc{V-Var}.  Since $v'[x \Leftarrow v] = v$, we can
    conclude $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.

    Otherwise, we are in the case where $v'$ is some variable $y \neq x$.
    Then, given that $v'[x \Leftarrow v] = v' = y$ and since $\Gamma(y) =
    \Gamma[x \mapsto s]$, we can see that $\typeRule{\Gamma[x \mapsto
    s]}{v'}{t}$ implies that $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.
  \end{case}

  \begin{case}[\textsc{V-Abs}]
    The premise in this case becomes $\typeRule{pc, \Gamma[x \mapsto s][y
    \mapsto t'][f \mapsto (t' \to t'')^l]}{e'}{t''}$.

    Let us assume that $x$, $f$, and $y$ are distinct, since otherwise this
    would reduce to a similar situation to that of \textsc{V-Var} when the
    variable substitution leaves $v'$ unchanged, which is trivial.  Since they
    are distinct, $\Gamma[x \mapsto s][y \mapsto t'][f \mapsto (t' \to t'')^l]$
    is equivalent to $\Gamma[y \mapsto t'][f \mapsto (t' \to t'')^l][x \mapsto
    s]$ and so we can apply the inductive hypothesis and \textsc{V-Abs} to show
    $\typeRule{\Gamma}{v'[x \Leftarrow v]}{t}$.
  \end{case}

  \begin{case}[\textsc{V-Bracket}]
    In this case the premises are of the form $\typeRule{\Gamma[x \mapsto
    s]}{v'_i}{t}$.  By Lemma~\ref{lem:projection}, we know that
    $\typeRule{\Gamma}{v}{s}$ implies $\typeRule{\Gamma}{\lift{v}_i}{s}$.  By
    the inductive hypothesis, we can then say that $\typeRule{\Gamma}{v'_i[x
    \Leftarrow v]}{t}$ holds.  Thus, we can apply \textsc{V-Bracket} to obtain
    $\typeRule{\Gamma}{<v'_1[x \Leftarrow v] | v'_2[x \Leftarrow v]>}{t}$,
    proving the statement.
  \end{case}

  \begin{case}[\textsc{V-Sub}, \textsc{E-Value}, \textsc{E-App}, \textsc{E-If},
    \textsc{E-Let}, \textsc{E-Sub}]
    Applying the inductive hypothesis on the premises and applying the
    associated rule proves these cases.
  \end{case}

  \begin{case}[\textsc{E-Bracket}]
    This case is similar to the case of \textsc{V-Bracket}.
    Lemma~\ref{lem:guard} allows us to apply \textsc{E-Bracket} where necessary.
  \end{case}
\end{proof}

@@

We would like to thank Jon Sailor for chatting with us on the phone to help work
through some of the more difficult problems we ran into using Coq.
